---
title: "calculate scores using prepared data layers"
output: 
  html_document:
    self_contained: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
  
```


*TODO @jules32: some kind of overview like:* This chapter will take you from prepared data to calculated scores. It requires book keeping, `R` functions, and ultimately, the `R` package `ohicore`...

- *In this example, we will substitute a single global data layer with local data and recalculate scores without modifying the goal model itself. The \@ref(update) chapter will show an example for updating goal models.* 
- *screenshot of RStudio window, overview of `configure_toolbox.r` and `calculate_scores.r`, etc* 
- *bigger overview of how things work together and workflow: functions.r is a big list of goal models, as R functions. You can run all of them at the once or individually. In this chapter we will work on just one goal model*

----

TODO @ningningj and @jules32: we need to be more explicit about the overview here. 
Eg: In this chapter we're going to 
    - single goal model with updated data, basically line-by-line to understand the architecture of goal models (access the correct data layers, calc status, calc trend, make sure the final output is standard for what the `scores variables` need. 
    - then calculate all goal models and scores, including pressures, and resilience. 

- will probably need a section "what is ohicore and how does it operate with my ohi+ repo?"
- introduce which data layer we modified and which goal model we are using as an example (eg AO)
    

---- 

In this section, we won't be modifying goal models yet. Instead, we will get you familiar with the construct of `functions.R` and the sequence of events required to calculate scores. This means we'll take the data layer that you just prepared \@ref(prep) and follow these steps to calculate scores: 

1. Confirm that data layers are saved and registered **TODO @ningningj: saving and registering should probably be in the \@ref(prep) chapter?** We could save it in a different .Rmd for now and have it be a subchapter of the prep chapter, referencing it here?: 

1. Load `ohicore` and check data layers _in `configure_toolbox.r`_

1. Calculate goal scores _in `conf/functions.r`_ **TODO @ningningj: let's not say 'update' in this chapter, let's save that for the following chapter. Here we just want to calculate goal scores** 
  - load data (`functions.r`)
  - calculate status scores (`functions.r`)
  - calculate trend scores (`functions.r`)
  - combine status and trend scores (`functions.r`)

4. check and possibly update how the function is called _in `conf/goals.csv`_

5. calculate the remaining component of OHI scores _in `calculate_scores.r`_

We will go through each of the scripts mentioned above. Here is a general structure to show where those files are in relation to the entire calculation process: 

**TODO @ningningj: this figure may be more clear here if the red boxes are removed and instead placed around the green top pieces "modify goal models" and "calculate scores"?**
![](https://docs.google.com/drawings/d/1zHe1Gp2L7xN04w3NO_uQvOrl47Ug_oogyj2E338J90U/pub?w=960&h=600)

Now let's get started and give it a try! 

## Step 1: Save and register data layers TODO @ningingj: let's move this into a separate prep Rmd

Before processed data can be used by the Toolbox for calculations, they need to be saved and registered here: 

![](https://docs.google.com/drawings/d/1IwHf8pMBlpzOIYKL0hoEF5rETGA765Qtwkz7es7B3Rg/pub?h=500)

### 1.1 Save data layers in the *layers* folder

When modifying existing or creating new data layers in the prep folder, save the ready-to-use layers in this folder. We recommend saving it as a new *.csv* file with: 

- a prefix identifying the goal (eg. `fis_`)
- a suffix identifying your assessment (eg. `_sc2016.csv`). 

Modifying the layer name provides an easy way to track which data layers have been updated regionally, and which rely on global data. Then, the original layers (`_gl2016.csv` and `_gl2016placeholder.csv`) can be deleted.  

_**Tip**: filenames should not have any spaces: use an underscore instead. This will reduce problems when R reads the files._

### 1.2 Register data layers in `layers.csv`  

After you saved the data layer and named it properly, now you can register it in `layers.csv`. This will tell the Toolbox where to look for data when you calculate scores.  

![](https://docs.google.com/drawings/d/1juueRiA0gevOoHH_8owzLNanyQ-H1mAOEl7L_l5YSJ0/pub?h=600)  

_If a layer simply has a new filename, only the *filename* column needs to be updated._


## Source `configure_toolbox.r`
**note @ningningj: bookdown will take care of numbering so we can remove the 'Steps' I think 

Now you have fully prepared and registered the data with the Toolbox, there is just one more step before calculations! We have to make sure everything is properly configured for the Toolbox: 

**TODO @ningningj, highlight `configure_toolbox.r` in this figure instead of functions.r**

![](https://docs.google.com/drawings/d/1CKNjIORLJOEaV2XW4Z9QuFLN7q1lJ1PaEGg2dSG1-_o/pub?h=450)

This script combines everything required to calculate OHI scores and checks that they are properly formatted, and will minimize potential errors down the line. It makes sure that your data and goal models are ready to be used to calculate scores. 
**TODO @ningningj/@jules32: add section/language about how to interpret errors: errors are identifying YOUR problems, not ohicore's (said in a nicer way)**

More specifically, sourcing `configure_toolbox.r` will: 

1. load the necessary `R` packages, including `ohicore`
2. load scenario configurations (eg. functions.r) TODO @ningningj: I think we need to write out what `ohicore::Conf('conf')` etc are doing, this is too vague
3. check that data layers are properly formatted and registered 
4. load data layers so that you can call them in functions.r 

_**Note**: repeat this step whenever you save a new data layer or change a goal function._ **TODO @ningningj: this is a really important point, should not be a note but a statement, and probably in the bigger overview above

_**Tip**: "Source" to run all the lines_

![](https://docs.google.com/drawings/d/1JvSJKyiwvp92--obTwHT3IYht1XE0823sORJ4FdnX74/pub?h=450)

## Explore goal model
**TODO @ningningj: let's not call this update, that's for the next chapter. This is just making sure the appropriate data layer is called. 

Now you are ready to examine the goal model in `functions.r` and confirm that your data layer is appropriately called. `functions.r` is in the `conf` folder. TODO @ningningj/@jules32: more here

![](https://docs.google.com/drawings/d/1CKNjIORLJOEaV2XW4Z9QuFLN7q1lJ1PaEGg2dSG1-_o/pub?h=450)

TODO @ningningj/@jules32: develop paragraphs from these bullets: 
- `functions.r` has goal models (status and trend) for each goal in the assessment. Each goal model is a separte R function. Quick overview of how an R function is structured
- `R` code from the most recent OHI global assessment as a starting place. You can run through the reference script line-by-line to learn how it has been done. 

Open `functions.r`, go to the appropriate goal function. We will use Artisanal Fishing Opportunity (AO) as an example. 

_**Tip**: Clicking the bottom left corner of Console will show you a drop-down menu of all functions. It's a shortcut to jump to the appropriate section or goal model._

![](https://docs.google.com/drawings/d/1Z2AVtVfIZEd_5GDcmaLb8mRpErKTDzmcbg0SBMxKsxs/pub?h=450)

*Note @ningningj: this figure is too small to see, less white space would help*

_Each goal model is written as a function, and you will modify the goal model within the confines of each function._ **TODO @ningningj: see note above but this needs to be more prominent**

### SelectLayersData

Run this chunk of code and the data layers will be loaded, joined, and ready to be manipulated further: 

**TODO @ningningj/@jules32: SelectLayersData is from ohicore, what it's doing...**

```{r load data}

## "SelectLayersData" is an ohicore funtion to call the appropriate data layer by its layer name registered in`layers.csv` (eg. "ao_access")

## "select"" is a function from the dplyr package to let you select only the columns you would need

r <- SelectLayersData(layers, layers = 'ao_access', narrow=TRUE) %>%
    select(region_id=id_num, access=val_num)
r <- na.omit(r)

ry <- SelectLayersData(layers, layers = 'ao_need', narrow=TRUE) %>%
    select(region_id = id_num, year, need=val_num) %>%
    left_join(r, by="region_id")

```


> Tip: it's always a good idea to check what your data looks like and make sure there are no glaring errors. A couple of commonly used functions are: head() and str()

### 3.2 Calculate status 

Once you make sure your raw data have been loaded and joined in the right format, you can start calculate status scores!

**TODO @ningningj/@jules32: this syntax uses the `tidyverse` and chaining...look xxx to learn more** 

```{r}
# model: this step calculates status scores of all years, using the goal model. 

## "mutate" is another commonly used function from dplyr that allows you to add a new column to the data frame
## Note that "Sustainability" and "status_year" have been defined at the start of the AO function 
  Sustainability=1.0
  status_year = 2015

  ry <- ry %>%
    mutate(Du = (1 - need) * (1 - access)) %>%
    mutate(status = (1 - Du) * Sustainability)

# status: status scores are typically the most recent year of all the years you have calculated. 

  r.status <- ry %>%
    filter(year==status_year) %>%
    select(region_id, status) %>%
    mutate(status=status*100)
```

### 3.3 Calculate trend 

Trend scores are typically based on linear regression of status scores from the most recent five years. 

```{r}
 # choose trend years (eg. most recent five years)

  trend_years <- (status_year-4):(status_year)
  adj_trend_year <- min(trend_years)

  r.trend = ry %>%
    group_by(region_id) %>% 
    # linear model: 
    do(mdl = lm(status ~ year, data=., subset=year %in% trend_years),
       adjust_trend = .$status[.$year == adj_trend_year]) %>%
    # extract the coefficient of year and produce a trend score
    summarize(region_id, trend = ifelse(coef(mdl)['year']==0, 0, coef(mdl)['year']/adjust_trend * 5)) %>%
    # make sure that the scores are between -1 and 1
    mutate(trend = ifelse(trend>1, 1, trend)) %>%
    mutate(trend = ifelse(trend<(-1), (-1), trend)) %>%
    mutate(trend = round(trend, 4))

```

### 3.4 Combine status and trend scores 

Choose only region_id and score, and add two more columns identifying score dimension (status or trend) and goal name. 
```{r}
 scores = r.status %>%
    select(region_id, score=status) %>%
    mutate(dimension='status') %>%
    rbind(
      r.trend %>%
        select(region_id, score=trend) %>%
        mutate(dimension='trend')) %>%
    mutate(goal='AO')
```

## Step 4. Update goal call in `goals.csv`**

`goals.csv` in the `conf` folder provides input information for `functions.r`, particularly about function calls. These are indicated by two columns: *preindex_function* (functions for all goals that do not have sub-goals, and functions for all sub-goals) and *postindex_function* (functions for goals with sub-goals).

In the `preindex_fuction`, you could specify variables such as _status_year_ and _trend_year_, which you can call in your goal function. Note that it is not necessary to specify those variables. If you do not use them in your function as in the CS example, you could delete those variables in `preindex_function`.

_Changing goal weights will be done here by editing the value in the *weight* column. Weights do not need to be 0-1 or add up to 10; weights will be scaled as a proportion of the number of goals assessed. The ten goals are weighted equally by default._

![](https://docs.google.com/drawings/d/17BgYSw2sHbZvHNjUqBlTG-kCOAAn7o6a65O37s0S_es/pub?h=500)

![](https://docs.google.com/drawings/d/1o2wtJ9KCPDyGPH9Y4unmALG6BlxX9lmJ_PakDDiQrLo/pub?h=500)

## Step 5. Calculate overall scores 
**TODO @ningningj/@jules32: I think we'll want to make this a distinct thing that also includes pressures/resilience, etc, whereas they've just been walking through the status/trend for a single goal model above. Does that make sense? We can think of the organization...**

Hooray you've modified a goal function! Now let's calculate the remaining score components (pressures, resilience, likely future state, and overall score). This is done in `calculate_scores.r`

_We'll learn how to modify Pressures and Resilience matrix in a different chapter._ 


```{r}
# calculate scenario scores
scores = CalculateAll(conf, layers, debug=F)

# save scores as .csv file, tables and figures
write.csv(scores, 'scores.csv', na='', row.names=F)
```

After the calculation is done, you should be able to see the compiled score sheet for all goals in all regions in sub-country/scores.csv.

It is very likely that during the CalculateAll process you’ll encounter problems and see error messages. In most cases, the error messages can specify what the error is and in which step it occurs, which should be helpful for trouble shooting. Some commonly occurring errors and how to fix them can be found in the Troubleshooting section of the manual.

_**Tip**: whenever there are changes made (addition, deletion, changes in .csv or r script), you would be notified in the Git window. So here, since you added a new data layer and/more modified a function, and are expecting changes in score.csv, you should see it in Git. This is a good place to double check whether you have made the right changes._
